---
title: "CommonMind Consortium: DLPFC"
subtitle: "Analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
---


<!---

cd /hpc/users/hoffmg01/work/covarr-de
# rm -rf combined_analysis/diff.networks_cache
ml git pandoc  udunits proj gdal geos
git pull
R

system("git pull")


rmarkdown::render("combined_analysis/diff.networks.Rmd")


--->

Since we are generating networks across cohorts and sexes, identifiying conserved clusters using standard methods becomes non-trivial.  Instead of defining clusters for each subset and evaluating the difference, we first compute the differences in the correlation networks, then identify a single set a clusters based on this meta-network.  Then use sLED test on difference network, or use boxM_permute on original residuals, but with consensus clusters.

But if concensus models are constructed to be differential by definition, then these clusters could show differential correlation even under the null.  Therefore, use set A as discovery for M/F differential network, and replicated in set B.



```{r setup, include=FALSE}
suppressPackageStartupMessages({
library(synapser)

synLogin()

library(WGCNA)
library(variancePartition)
library(decorate)
library(ggrepel)
library(cowplot) 
library(plyr) 
library(qvalue)
}) 
source("../common_functions/common_functions.R")   
# source("common_functions/common_functions.R")   
  
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

```{r load.data}
resid.add = readRDS(synGet('syn22416489')$path)
METADATA = readRDS(synGet('syn22416411')$path)
METADATA$Reported_Gender = factor(METADATA$Reported_Gender, c("Male", "Female"))
```


## Difference networks
```{r diff.networks}

# Compute difference networks
##############################

resVersion = "Sex"
variable = 'Reported_Gender'

diff.net = get_difference_network(resid.add[[resVersion]], METADATA, variable)
   
# Discover clusters in first cohort
#----------------------------------
C.diff.discovery = diff.net[[1]]

hcl = hclust( as.dist(C.diff.discovery) )

dynamicMods = cutreeDynamic(hcl, distM=C.diff.discovery, minClusterSize = 20)

dynamicColors = labels2colors(dynamicMods, colorSeq = colors())
tab = table(dynamicColors)
length(tab)
```

```{r testing}

# Hypothesis testing
#-------------------

df_test = test_differential_correlation(resid.add[[resVersion]], C.diff.discovery, dynamicColors, METADATA)  
```

```{r plot.discovery}
# plot discovery and testing
ggplot(df_test, aes(Module, -log10(FDR))) + geom_bar(stat='identity', fill=df_test$Module) + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) + coord_flip() + facet_wrap(~Cohort) + ylab(bquote(-log[10]~FDR)) + scale_y_continuous(expand=c(0,0)) + geom_hline(yintercept = -log10(0.05), color="red", linetype="dashed")
```

```{r modules.in.discovery, fig.width=8, fig.height=20}

figList = lapply(df_test$Module[1:5], function(mod){

	plot_module( mod, df_test, METADATA, dynamicColors, C.diff.discovery, resid.add[[resVersion]], variable, "MSSM-Penn-Pitt")  
	})
  
plot_grid( plotlist = figList, ncol=1)
```




```{r second.cohort}
df_cohort2 = df_test[df_test$Cohort == "NIMH-HBCC",]
df_cohort2$Module = factor(df_cohort2$Module, unique(df_cohort2$Module))
table(df_cohort2$FDR < 0.05)

df = df_cohort2[df_cohort2$P.Value < 0.1,]

ggplot(df, aes(Module, -log10(FDR))) + geom_bar(stat='identity', fill=df$Module) + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) + coord_flip() + facet_wrap(~Cohort) + ylab(bquote(-log[10]~FDR)) + scale_y_continuous(expand=c(0,0)) + geom_hline(yintercept = -log10(0.05), color="red", linetype="dashed")
```

```{r modules.in.second.cohort, fig.width=8, fig.height=40}

i = which(df_cohort2$FDR < 0.05)[1:5]

figList = lapply(df_cohort2$Module[i], function(mod){

	plot_module( mod, df_test, METADATA, dynamicColors, C.diff.discovery, resid.add[[resVersion]], variable, "NIMH-HBCC")  
	})
  
plot_grid( plotlist = figList, ncol=1)
```

<!---


# need to generate correct residuals

GWAS hits, TWAS, SCHEMA
ChrX and ChrY genes with autism and Scz

DE within each sex, on chrX and chrY.
	drop chrY genes for computing FDR?

Can we predict sex given autosomal genes?

PMID: 26892004
https://science.sciencemag.org/content/369/6509/eaba3066
https://www.nature.com/articles/s41576-018-0083-1


test: how much sex differences are given by age, cell type and institution.	






library(mvtnorm)
library(Rfast)

X = matrnorm(300, 1000)
Y = matrnorm(300, 1000)

C_x = cor(X)
C_y = cor(Y)

df_cor = data.frame( 	a = C_x[lower.tri(C_x)],
						b = C_y[lower.tri(C_y)])


with(df_cor, plot(a,b))

a = calc_sde(id=1,points=df_cor);

Sigma = cov(df_cor)

--->
